\Rpackage{MSnbase}: labelled MS2 data pre-processing, visualisation and quantification.

### 2012-02-03 inst/doc/MSnbase-demo/MSnbase-demo.Rnw
Update vignette with 'combine' section + minor fixes.
CHANGES IN VERSION 1.3.9
-------------------------
 o updated demo vignette <2012-02-03 Fri>
 o sorting numeric subsets in "[" pSet, as unsorted 
   numerical indexes fails  <2012-02-03 Fri>
 o Added match.arg in combineFeatures so that 
   a unique default value (the first) is used when no 
   fun is specified <2012-02-03 Fri>




git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MSnbase@62533 bc3139a8-67e5-0310-9ffc-ced21a209358

\Rpackage{MSnbase} aims are providing a reproducible research framework to proteomics data analysis. 
It should allow researcher to easily mine mass spectrometry data, explore the data and its statistical properties 
and visually display these.

@@

<<readdata, echo=TRUE, cache=FALSE>>=
file <- dir(system.file(package="MSnbase", dir="extdata"),
            full.names=TRUE, pattern="mzXML$")
rawdata <- readMSData(file, msLevel=2, verbose=FALSE)
@

@@

\begin{figure}[!ht]
<<spectrum-plot,fig=T,echo=T,tikz=T,external=F,width=5,height=5>>=
plot(sp,reporters=iTRAQ4,full=TRUE) 
@ 
\caption{Raw MS2 spectrum with details about reporter ions.}
\label{fig:spectrum-plot}
\end{figure}

<<bsasel,echo=FALSE>>=
bsasel <- fData(itraqdata)$ProteinAccession=="BSA"
@

@@

<<subset,echo=TRUE>>=
sel <- fData(itraqdata)$ProteinAccession=="BSA"
bsa <- itraqdata[sel]
bsa
as.character(fData(bsa)$ProteinAccession)
@

@@

\begin{figure}[p]
<<msnexp-plot,fig=T,echo=T,tikz=T,external=F,width=5,height=6>>=
plot(bsa,reporters=iTRAQ4,full=FALSE)
@ 
\caption{Experiment-wide raw MS2 spectra. The y-axes of the individual 
  spectra are automatically rescaled to the same range. 
  See section \ref{sec:norm} to rescale peaks identically. }
\label{fig:msnexp-plot}
\end{figure}

@@

<<removePeaks,echo=TRUE,cache=FALSE>>=
experiment <- removePeaks(itraqdata,t=400,verbose=FALSE)
## total ion current
tic(itraqdata[["X55"]])
tic(experiment[["X55"]])
@

<<prepare-spectrum-plot2,echo=FALSE>>=
p1 <- plot(itraqdata[["X55"]],full=TRUE) + theme_gray(5) 
p2 <- plot(experiment[["X55"]],full=TRUE) + theme_gray(5) 
@

@@

<<clean,echo=TRUE,cache=FALSE>>=
## number of peaks
peaksCount(itraqdata[["X55"]])
peaksCount(experiment[["X55"]])
experiment <- clean(experiment,verbose=FALSE)
peaksCount(experiment[["X55"]])
@

@@

<<trimMz,echo=TRUE,cache=FALSE>>=
range(mz(itraqdata[["X55"]]))
experiment <- trimMz(experiment,mzlim=c(112,120))
range(mz(experiment[["X55"]]))
experiment
@ 
As can be seen above, all processing performed on the experiment is recorded 
and displayed as integral part of the experiment object.

@@

<<quantify,echo=TRUE,cache=FALSE,keep.source=T>>=
qnt <- quantify(experiment,
                method="trap",
                reporters=iTRAQ4,
                strict=FALSE,
                verbose=FALSE)
qnt
head(exprs(qnt))
@

@@

<<removeNa,echo=TRUE>>=
sum(is.na(exprs(qnt)))
whichRow <- which(is.na(exprs(qnt))) %% nrow(qnt)
qnt <- qnt[-whichRow,]
sum(is.na(exprs(qnt)))
@

@@

<<pheplus1, echo=TRUE, cache=FALSE>>=
library(Rdisop)
## Phenylalanine immonium ion
Fim <- getMolecule("C8H10N")
getMass(Fim)
isotopes <- getIsotope(Fim)
F1 <- isotopes[2,2]
F1
@

@@

<<purityCorrect,echo=TRUE,cache=FALSE>>=
impurities <- matrix(c(0.929,0.059,0.002,0.000,
                       0.020,0.923,0.056,0.001,
                       0.000,0.030,0.924,0.045,
                       0.000,0.001,0.040,0.923),
                     nrow=4)
qnt.crct <- purityCorrect(qnt,impurities)
head(exprs(qnt))
head(exprs(qnt.crct))
@

@@

<<normalise,echo=TRUE,cache=FALSE>>=
qnt.max <- normalise(qnt,"max")
qnt.sum <- normalise(qnt,"sum")
qnt.quant <- normalise(qnt,"quantiles")
qnt.qrob <- normalise(qnt,"quantiles.robust")
qnt.vsn <- normalise(qnt,"vsn")
@

@@

<<prepare-msnset-nrom-plot,cache=FALSE,echo=FALSE>>=
p <- plot(normalise(experiment[bsasel],"max"),reporters=iTRAQ4,full=FALSE)
p <- p + theme_gray(7)
@

@@

<<combineFeatures,echo=TRUE,cache=FALSE>>=
qnt2 <- combineFeatures(qnt,groupBy=gb,fun="median")
qnt2
@

@@

<<incompdiss,echo=TRUE,cache=FALSE,keep.source=T>>=
iTRAQ5
incompdiss <- quantify(itraqdata,
                       method="trap",
                       reporters=iTRAQ5,
                       strict=FALSE,
                       verbose=FALSE)
head(exprs(incompdiss))
@

@@

\begin{figure}[!hbt]
  \begin{center}
<<incompdiss-plot,fig=T,echo=F,tikz=T,external=F,width=4,height=3.5>>=
dfr <- melt(exprs(incompdiss))
colnames(dfr) <- c("feature","reporters","intensity")
qplot(x=reporters,y=intensity,data=dfr,geom="boxplot") + theme_gray(8)
@ 
\caption{ Boxplot comparing intensities of the 4 reporter ions an the 
  incomplete dissociation specific peak. }
\label{fig:incompdiss}
  \end{center}
\end{figure}

@@

>>>>

\Rpackage{MSnbase} \citep{Gatto2012} aims are providing a reproducible research framework to proteomics data analysis. 
It should allow researcher to easily mine mass spectrometry data, explore the data and its statistical properties 
and visually display these.

@@

<<readdata, echo=TRUE, cache=FALSE>>=
file <- dir(system.file(package = "MSnbase", dir = "extdata"),
            full.names = TRUE, pattern = "mzXML$")
rawdata <- readMSData(file, msLevel = 2, verbose = FALSE)
@

@@

\begin{figure}[!ht]
<<spectrum-plot,fig=T,echo=T,tikz=T,external=F,width=5,height=5>>=
plot(sp, reporters = iTRAQ4, full = TRUE) 
@ 
\caption{Raw MS2 spectrum with details about reporter ions.}
\label{fig:spectrum-plot}
\end{figure}

<<bsasel,echo=FALSE>>=
bsasel <- fData(itraqdata)$ProteinAccession == "BSA"
@

@@

<<subset,echo=TRUE>>=
sel <- fData(itraqdata)$ProteinAccession == "BSA"
bsa <- itraqdata[sel]
bsa
as.character(fData(bsa)$ProteinAccession)
@

@@

\begin{figure}[p]
<<msnexp-plot,fig=T,echo=T,tikz=T,external=F,width=5,height=6>>=
plot(bsa, reporters = iTRAQ4, full = FALSE)
@ 
\caption{Experiment-wide raw MS2 spectra. The y-axes of the individual 
  spectra are automatically rescaled to the same range. 
  See section \ref{sec:norm} to rescale peaks identically. }
\label{fig:msnexp-plot}
\end{figure}

@@

<<removePeaks,echo=TRUE,cache=FALSE>>=
experiment <- removePeaks(itraqdata, t = 400, verbose = FALSE)
## total ion current
tic(itraqdata[["X55"]])
tic(experiment[["X55"]])
@

<<prepare-spectrum-plot2,echo=FALSE>>=
p1 <- plot(itraqdata[["X55"]], full = TRUE) + theme_gray(5) 
p2 <- plot(experiment[["X55"]], full = TRUE) + theme_gray(5) 
@

@@

<<clean,echo=TRUE,cache=FALSE>>=
## number of peaks
peaksCount(itraqdata[["X55"]])
peaksCount(experiment[["X55"]])
experiment <- clean(experiment, verbose = FALSE)
peaksCount(experiment[["X55"]])
@

@@

<<trimMz,echo=TRUE,cache=FALSE>>=
range(mz(itraqdata[["X55"]]))
experiment <- trimMz(experiment, mzlim = c(112,120))
range(mz(experiment[["X55"]]))
experiment
@ 
As can be seen above, all processing performed on the experiment is recorded 
and displayed as integral part of the experiment object.

@@

<<quantify,echo=TRUE,cache=FALSE,keep.source=T>>=
qnt <- quantify(experiment,
                method = "trap",
                reporters = iTRAQ4,
                strict = FALSE,
                verbose = FALSE)
qnt
head(exprs(qnt))
@

@@

<<removeNa,echo=TRUE>>=
sum(is.na(exprs(qnt)))
whichRow <- which(is.na(exprs(qnt))) %% nrow(qnt)
qnt <- qnt[-whichRow, ]
sum(is.na(exprs(qnt)))
@

@@

<<pheplus1, echo=TRUE, cache=FALSE>>=
library(Rdisop)
## Phenylalanine immonium ion
Fim <- getMolecule("C8H10N")
getMass(Fim)
isotopes <- getIsotope(Fim)
F1 <- isotopes[2, 2]
F1
@

@@

<<purityCorrect,echo=TRUE,cache=FALSE>>=
impurities <- matrix(c(0.929, 0.059, 0.002, 0.000,
                       0.020, 0.923, 0.056, 0.001,
                       0.000, 0.030, 0.924, 0.045,
                       0.000, 0.001, 0.040, 0.923),
                     nrow = 4)
qnt.crct <- purityCorrect(qnt, impurities)
head(exprs(qnt))
head(exprs(qnt.crct))
@

@@

<<normalise,echo=TRUE,cache=FALSE>>=
qnt.max <- normalise(qnt, "max")
qnt.sum <- normalise(qnt, "sum")
qnt.quant <- normalise(qnt, "quantiles")
qnt.qrob <- normalise(qnt, "quantiles.robust")
qnt.vsn <- normalise(qnt, "vsn")
@

@@

<<prepare-msnset-nrom-plot,cache=FALSE,echo=FALSE>>=
p <- plot(normalise(experiment[bsasel], "max"),
          reporters = iTRAQ4, full = FALSE)
p <- p + theme_gray(7)
@

@@

<<combineFeatures,echo=TRUE,cache=FALSE>>=
qnt2 <- combineFeatures(qnt, groupBy = gb, fun = "median")
qnt2
@

@@

<<incompdiss,echo=TRUE,cache=FALSE,keep.source=T>>=
iTRAQ5
incompdiss <- quantify(itraqdata,
                       method = "trap",
                       reporters = iTRAQ5,
                       strict = FALSE,
                       verbose = FALSE)
head(exprs(incompdiss))
@

@@

\begin{figure}[!hbt]
  \begin{center}
<<incompdiss-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=2.5>>=
dfr <- melt(exprs(incompdiss))
colnames(dfr) <- c("feature", "reporters", "intensity")
dfr$reporters <- sub("iTRAQ5.", "", dfr$reporters)
repsum <- apply(exprs(incompdiss), 1, function(x) sum(x[1:4]))
dfr2 <- data.frame(iTRAQ1to4 = repsum, iTRAQ5 = exprs(incompdiss)[,5])
p1 <- ggplot(data = dfr, aes(x = reporters,y = log10(intensity))) +
  geom_boxplot() +
  theme_gray(6)
p2 <- ggplot(data = dfr2, aes(x = log10(iTRAQ1to4), y = log10(iTRAQ5))) +
  geom_point(alpha = I(1/2)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  stat_smooth(method = lm, se = FALSE) +
  xlab(label = expression(log[10]~sum~114~to~117)) +
  ylab(label = expression(log[10]~145)) +
  theme_gray(6)
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
print(p1, vp = vplayout(1, 1))
print(p2, vp = vplayout(1, 2))
@ 
\caption{ Boxplot and scatterplot comparing intensities of the 4 reporter ions 
  (or their sum, on the right) and the incomplete dissociation specific peak. }
\label{fig:incompdiss}
  \end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Combining \Robject{MSnSet} instances}\label{sec:combine}

Default MSnSet feature names (\Robject{X1}, \Robject{X2}, \ldots) and sample names 
(\Robject{iTRAQ4.114}, \Robject{iTRAQ4.115}, \Robject{iTRAQ4.116}, \ldots) are not very informative.
The features and samples of these anonymous quantitative data-sets should be updated prior 
to additional downstream analysis.

Lets create two \Robject{MSnSet}s from the same raw data to simulate two different 
independent experiments. As done previously (see section \ref{sec:feataggregation}), 
we combine the specta based on the proteins they have been identified to belong to.
Features can thus naturally be named using protein accession numbers.
Alternatively, if peptide sequences would have been used as grouping factor in 
\Rfunction{combineFeatures}, then these would be good feature name candidates.

<<make2exps, echo=TRUE, cache=FALSE>>=
set.seed(1)
i <- sample(length(itraqdata), 35)
j <- sample(length(itraqdata), 35)
exp1 <- quantify(itraqdata[i], reporters = iTRAQ4, verbose = FALSE)
exp2 <- quantify(itraqdata[j], reporters = iTRAQ4, verbose = FALSE)
table(featureNames(exp1) %in% featureNames(exp2))
exp1 <- combineFeatures(exp1, groupBy = fData(exp1)$ProteinAccession)
exp2 <- combineFeatures(exp2, groupBy = fData(exp2)$ProteinAccession)
head(featureNames(exp1))
featureNames(exp1) <- fData(exp1)$ProteinAccession
featureNames(exp2) <- fData(exp2)$ProteinAccession
head(featureNames(exp1))
@

Setting the feature names is critical when multiple experiments are to be combined, as 
this is done using common features as defined by their names (see below).

Sample names should also be updated to replace reporter names with relevant identifiers; 
the individual reporter data is stored in the \Robject{phenoData} and is not lost.
A convenience function \Rfunction{updateSampleNames} is provided to append the \Robject{MSnSet}'s 
variable name to the already defined names, although in general, biologically relevant 
identifiers are preferred.

<<renameSamples, echo=TRUE, cache=FALSE>>=
sampleNames(exp1)
exp1 <- updateSampleNames(exp1)
sampleNames(exp1)
sampleNames(exp1) <- c("Ctrl1", "Cond1", "Ctrl2", "Cond2")
sampleNames(exp2) <- c("Ctrl3", "Cond3", "Ctrl4", "Cond4")
@

At this stage, it is not yet possible to combine the two experiments, 
because their feature data is not compatible; they share the same feature 
variable labels, i.e. the feature data column names 
(\Sexpr{paste(head(fvarLabels(exp1), n = 3), collapse=", ")}, \ldots), 
but the content is different because the original data was, in particular 
all the spectrum data. Feature data with identical labels (columns in the data frame) 
and names (row in the data frame) are expected to have the same data and produce an
error if not conform.

<<fdatanames, echo=TRUE, cache=FALSE>>=
all(fvarLabels(exp1) == fvarLabels(exp1))
fData(exp1)["BSA", 1:4]
fData(exp2)["BSA", 1:4]
@

Instead of removing these overlapping and different feature data columns, 
one can use a second convenience function, \Rfunction{updateFvarLabels}, 
to update feature labels based on the experiements variable name and maintain all the metadata.

<<renameFvars, echo=TRUE, cache=FALSE>>=
exp1 <- updateFvarLabels(exp1)
exp2 <- updateFvarLabels(exp2)
head(fvarLabels(exp1))
head(fvarLabels(exp2))
@

It is now possible to combine \Robject{exp1} and \Robject{exp2} with the 
\Rfunction{combine} method. The new experiment will contain the union of 
the feature names of the individual experiments with missing values 
inserted appropriately.

<<combine, echo=TRUE, cache=FALSE>>=
exp12 <- combine(exp1, exp2)
dim(exp12)
pData(exp12)
exprs(exp12)[25:28, ]
exp12
@

In summary, when experiments need to be combined, one needs to 
(1) clarify the sample names using \Rfunction{updateSampleNames} or,
better manually, for biological relevance and (2) update the feature 
data variable labels with \Rfunction{updateFvarLabels}. 
The individual experiments (there can be more than 2) can then easily be combined 
with the \Rfunction{combine} method while retaining the meta-data.

@@

% LocalWords:  renameSamples
